Type Classes for Mathematics in Type Theory

### 2010-06-30 jar.tex
Fix TODO naming author of unification-hints paper.
Polishing.

\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\lstset{literate=
  % symbols that actually occur as unicode in our source:
  {λ}{{$\uplambda\ $}}1
  {σ}{{$\upsigma$}}1
  {η}{{$\upeta$}}1
  {φ}{{$\upphi$}}1
  {∃}{{$\exists$}}1
  {→}{{$\to\ $}}1
  {≠}{{$\ne\ $}}1
  {¬}{{$\neg\ $}}1
  {⟶}{{$\longrightarrow\ $}}1
  {⇛}{{$\Rrightarrow\ $}}1
  {∧}{{$\land$}}1
  {∀}{{$\forall$}}1
  {Π}{{$\Uppi$}}1
  {η}{{$\upeta$}}1
  {⊓}{{$\sqcap$}}1
  {∘}{{$\circ\ $}}1
  {◎}{{$\odot$}}1
  { ≡ }{{$\equiv\ $}}1
  % things we can't make pretty in the actual source, but can make pretty here!:
  {=>}{{$\,\Rightarrow\ $}}1
  {==>}{{$\Rightarrow\ $}}1
  {<-}{{$\leftarrow\ $}}1
  {~}{{$\sim\ $}}1
}

@@

In the absence of type classes, these are all very real problems, and for this reason the two biggest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and Ssreflect~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.

@@

Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed and implemented in the Matita system to address exactly this problem. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution that avoids the problem altogether by using predicate-like type classes in place of bundled records.

@@

The observant reader may wonder whether it might be beneficial to go one step further and unbundle proof of laws and inherited substructures as well. This is not the case, because there is no point in sharing them. After all, by (weak) proof irrelevance, the ``value'' of such proofs can be of no consequence anyway. Indeed, parameterizing on proofs would be actively harmful because instantiations differing only in the proof argument would express the same thing yet be nonconvertible, requiring awkward conversions and hindering automation.

@@

Without further support from the system, this would have to be written as
\begin{lstlisting}
  Theorem L G e op {P: SemiGroup G e op}:
    Π x y z, e (op (op x y) z) (op (op z y) x).
\end{lstlisting}
(The curly brackets mark \lstinline|P| as an implicit argument.) Because \lstinline|e| and \lstinline{op} are freshly introduced local names, we cannot bind notations to them prior to this theorem. Hence, if we want notations, what we really need are canonical names for these components. This is easily accomplished with single-field type classes containing one component each, which we will call \emph{operational type classes}\footnote{These single-field type classes are used in the same way in the \lstinline|Clean| standard library~\cite{clean}.}:
\begin{lstlisting}
  Class Equiv A := equiv: relation A.
  Class SemiGroupOp A := sg_op: A $\to$ A $\to$ A.

@@

\begin{lstlisting}
  Theorem L G e op {P: SemiGroup G e op}: Π x y z,
    equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op x y) z).
\end{lstlisting}
where e and the op's are filled in by instance resolution.

@@

While we are on the topic of implicit generalization, we mention one inadequacy concerning their current implementation that we feel should be addressed for the facility to be a completely satisfying solution. While the syntax already supports variants (not shown above) that differ in how exactly different kinds of arguments are inferred and/or generalized, there is no support to have an argument be ``inferred if possible, generalized otherwise". We have found that the need for such a policy arises naturally when declaring a parameter of class type in a contexts where \emph{some} of its components are already available, while others are to be newly introduced. The current workaround in these cases involves providing names for components that are then never referred to, which is a bit awkward.

@@

Before we proceed with a discussion of the algebraic hierarchy based on type classes, we briefly highlight one specific operational type class because we will use it later and because it is a particularly nice illustration of another neat application of operational type classes. The operation in question is that of deciding a proposition:
\begin{lstlisting}
  Class Decision (P: Prop): Type := decide: sumbool P (¬ P).
\end{lstlisting}
Here, \lstinline|sumbool| is just the (informative) sum of proofs.

@@

Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive). In our hierarchy, these ``substructures'' by themselves establish the inheritance diagram in Figure~\ref{inheritancediagram}.

\begin{figure}
\centering
\includegraphics{hierarchy.pdf}
\caption{Inheritance relations established by \lstinline|:>| fields.}
\label{inheritancediagram}
\end{figure}

@@

Having described the basic principles of our approach, in the remainder of this paper we tour other parts of our development, further illustrating what a state of the art formal development of foundational mathematical structures can look like with a modern proof assistant.

@@

Initiality, too, is defined by a combination of an operational and a predicate class:
\begin{lstlisting}
  Context `{Category X}.
  Class InitialArrows (x: X): Type := initial_arrow: Π y, x ⟶ y.
  Class Initial (x: X) `{InitialArrows x}: Prop :=
    initial_arrow_unique: Π y (a: x ⟶ y), a = initial_arrow y.
\end{lstlisting}
The operational class \lstinline|InitialArrows| designates the arrows that originate from an initial object \lstinline|x| by virtue of it being initial. The \lstinline|Initial| class itself further requires these ``initial arrows'' to be unique. Having \lstinline|InitialArrows| as a type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where ``known to'' should be read as ``can be determined by instance resolution'').

@@

  Class Functor `(Fmap): Prop :=
    { functor_from: Category C
    ; functor_to: Category D
    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
      fmap (f ◎$$ g) = fmap f ◎$$ fmap g }.
\end{lstlisting}
We ought to say a few words about our use of \lstinline|fmap|. The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: ``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.

In Coq, for a term \lstinline|F| to function as though it had two different types simultaneously (namely the object map and the arrow map), there must either (1) be coercions from the type of F to either function, or (2) F must be (coercible to) a single function that is able to consume both object and arrow arguments. In addition to not being supported by Coq, option (1) would violate our policy of leaving components unbundled. For (2), if it could be made to work at all (which is not clear at all), F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map (namely ``\lstinline|Π x y, (x ⟶ y) → (F x ⟶ F y)|'') must refer to the object map.

We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two distinct and interdependent maps is an abuse of notation of sufficient severety to make it ill-suited to a formal development where software engineering concerns apply. Hence, we do not adopt this practice, and use ``\lstinline|fmap F|'' (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.

@@

To specify the natural numbers and the integers as initial objects in the categories of semirings and rings, respectively, definitions of these categories are needed. While one could define both of them ``manually'', greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, \emph{varieties} (that is, categories consisting of models for a fixed theory, with homomorphisms between them) of which can be defined generically.

@@

  Class AlgebraOps (σ: Signature) (carriers: sorts σ → Type)
    := algebra_op: Π o: operation σ, op_type carriers (σ o).
\end{lstlisting}
Because our carriers will normally be equipped with a setoid equality, we further define the predicate class \lstinline|Algebra|, stating that each of the operations respects the setoid equality on the carriers:
\begin{lstlisting}
  Class Algebra `{Π a, Equiv (carriers a)} `{AlgebraOps σ$$ carriers}: Prop :=
      { algebra_setoids:> Π a, Setoid (carriers a)
      ; algebra_propers:> Π o: σ, Proper (=) (algebra_op o) }.
\end{lstlisting}
The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an \lstinline|Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.

@@

To characterize structures such as semirings and rings, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
\begin{lstlisting}
Record EquationalTheory :=
  { et_sig:> Signature
  ; et_laws:> EqEntailment et_sig → Prop }.
\end{lstlisting}
An \lstinline|EqEntailment| consists of premises and a conclusion represented by an inductively defined statement grammar, which in turn uses an inductively defined term grammar. A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper,

We now introduce a predicate class designating algebras that satisfy the laws of an equational theory:
\begin{lstlisting}
Class InVariety
  (et: EquationalTheory) (carriers: sorts et → Type)
  {e: Π a, Equiv (carriers a)} `{!AlgebraOps et carriers}: Prop :=
  { variety_algebra:> Algebra et carriers
  ; variety_laws: Π s, et_laws et s → (Π vars, eval_stmt et vars s) }.
\end{lstlisting}

@@

In addition to this ``variety'' category, we also have categories of lawless algebras, as well as forgetful functors from the former to the latter, and from the latter to the category of setoids.

@@

To give a further taste of what universal algebra in our development looks like, we consider the definitions involved in the first homomorphism theorem~\cite{meinke1993universal} in more detail:
\begin{theorem}[First homomorphism theorem]
If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation defined by $a\sim b$ if and only if $f(a)=f(b)$ is a congruence on $A$, and the algebra $A/\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
\end{theorem}

A set of relations \lstinline|e| (one for each sort) is a congruence for an existing algebra if (1) \lstinline|e| respects that algebra's existing setoid equality, and (2) the operations with \lstinline|e| again form an algebra (namely the quotient algebra):
\begin{lstlisting}
Context `{Algebra σ$$ A}.
Class Congruence (e: Π s: sorts σ, relation (v s)): Prop :=
    { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
    ; congruence_quotient:> @Algebra et v e _ }.
\end{lstlisting}
We have proved that this natural and economical type-theoretic formulation that leverages our systematic integration of setoid equality is equivalent to the traditional definition of congruences as relations that, represented as sets of pairs, form a subalgebra of the product algebra.

@@

With these instances in place, we can now define the improved natural numbers specification:
\begin{lstlisting}
  Context `{SemiRing A} `{!NaturalsToSemiRing A}.
  Class Naturals: Prop :=
    { naturals_ring:> SemiRing A
    ; naturals_to_semiring_mor:> Π `{SemiRing B},
        SemiRing_Morphism (naturals_to_semiring A B)
    ; naturals_initial:> Initial (bundle_semiring A) }.
\end{lstlisting}
Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations. And this benefit is far from theoretical, as diverse representations of the natural numbers are abound; for instance, unary, binary, factor multisets, arrays of native machine words.

@@

To permit a \emph{generic} function operating on naturals to take advantage of specialized operations, we simply introduce an additional instance parameter:
\begin{lstlisting}
Definition calculate_things `{Naturals N} `{!Π n m: N, Decision (n = m)}
  (a b: nat): ... := ... decide (a = b) ... .
\end{lstlisting}
Without the \lstinline|Decision| parameter \lstinline|calculate_things| would be equally correct, but could be less efficient. Thus, by this scheme one can start by writing correct-but-possibly-inefficient programs that make use of generic operation instances, and then selectively improve efficiency of key algorithms simply by adding additional operational type class instance parameters where profiling shows it to make a significant difference, without changing their definition body.

@@

In our development we prove that the standard library's default rationals do indeed implement \lstinline|Rationals|, as do implementations of the \lstinline|QType| module interface. While the latter is rather ad-hoc from a theoretical perspective, it is nevertheless of great practical interest because it is used for the very efficient \lstinline|BigQ| rationals based on machine integers~\cite{machineintegers}. Hence, theory and programs developed on our \lstinline|Rationals| interface applies and can make immediate use of these efficient rationals. As stated earlier, we plan to rebase the computable real number implementation in CoRN on this interface, precisely so that it may be instantiated with efficient implementations like these.

We plan to provide an abstract interface for polynomials as a free commutative algebra. This would unify existing implementations such as coefficient lists and Bernstein polynomials.

@@

Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this ``quoting''. In~\cite{Hints}, TODO et. al. show that unification hints, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality by ingeniously using canonical structures.} This feature is absent from Coq, but fortunately type classes also allow us to do this, as we will now show.

@@

With these, we can define the primary ingredient, the \lstinline|Quote| class:
\begin{lstlisting}
  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V'): Type :=
    { quote: Expr (V + V')
    ; eval_quote: eval (merge l r) quote = n }.
\end{lstlisting}
We can think of \lstinline|Quote| as the type for a family of Prolog-like syntax-directed resolution functions, which will take as input \lstinline|V| and \lstinline|l| representing previously encountered holes (opaque subexpressions that could not be destructured further) and their values, along with a concrete term \lstinline|n| to be quoted. Their ``output'' will consist not only of the fields in the class, but also of \lstinline|V'| and \lstinline|r| representing additional holes and their values. Hence, a type class constraint of the form \lstinline|Quote x y z| should be read as ``quoting y with existing heap x generates new heap z''.

@@

Using a small additional utility lemma we can also easily quote equalities into \lstinline|eval p e = eval p' e'| form with a shared variable pack (so that identical expressions on both sides of the equation don't end up as different variables).

@@

Mathematically, finite sequences can be characterized as free monoids over sets. A categorical way of expressing this is in terms of adjunctions. As before, we \emph{could} fully embrace this perspective, paying no heed to practicality of implementation and usage, and define a relatively succinct type class for sequences as follows:
\begin{lstlisting}
Class PoshSequence
   (free: setoid.Object → monoid.Object) `{Fmap free}
   (singleton: id ⇛ monoid.forget ∘ free)
   (extend: `((x ⟶ monoid.forget y) → (free x ⟶ y))): Prop :=
   { sequence_adjunction: AltAdjunction singleton extend
   ; extend_morphism: `(Setoid_Morphism (extend x y)) }.
\end{lstlisting}
Here, \lstinline|monoid.forget| is the forgetful functor from monoids to sets.

@@

>>>>

\def\lstlanguagefiles{defManSSR.tex}
\lstset{language=SSR}
\lstset{literate=
  % symbols that actually occur as unicode in our source:
  {λ}{{$\uplambda\ $}}1
  {σ}{{$\upsigma$}}1
  {η}{{$\upeta$}}1
  {φ}{{$\upphi$}}1
  {∃}{{$\exists$}}1
  {→}{{$\to\ $}}1
  {≠}{{$\ne\ $}}1
  {¬}{{$\neg\ $}}1
  {⟶}{{$\longrightarrow\ $}}1
  {⇛}{{$\Rrightarrow\ $}}1
  {∧}{{$\land$}}1
  {∀}{{$\forall$}}1
  {Π}{{$\Uppi$ }}1
  {η}{{$\upeta$}}1
  {⊓}{{$\sqcap$}}1
  {∘}{{$\circ\ $}}1
  {◎}{{$\odot$}}1
  { ≡ }{{$\equiv\ $}}1
  % things we can't make pretty in the actual source, but can make pretty here!:
  {=>}{{$\,\Rightarrow\ $}}1
  {==>}{{$\Rightarrow\ $}}1
  {<-}{{$\leftarrow\ $}}1
  {~}{{$\sim$}}1
}

@@

In the absence of type classes, these are all very real problems, and for this reason the two largest formalizations of abstract algebraic structures in Coq today, CoRN~\cite{C-corn} and Ssreflect~\cite{Packed}, both use \emph{bundled} representation schemes, using records with one or more of the components as fields instead of parameters. For reflexive relations, the following is a fully bundled representation---the other end of the spectrum:
\begin{lstlisting}
  Record ReflexiveRelation: Type :=
    { rr_carrier: Type
    ; rr_rel: relation rr_carrier
    ; rr_proof: Π x, rr_rel x x }.
\end{lstlisting}
Superficially, this instantly solves the problems described above: reflexive relations can now be declared and passed as self-contained packages, and the \lstinline|rr_rel| projection now constitutes a ``canonical name'' for relations that are known to be reflexive, and we could bind a notation to it. While there is no conventional notation for reflexive relations, the situation is the same in the context of, say, a semiring, where we would bind ``$+$'' and ``$*$'' notations to the record field projections for the addition and multiplication operations, respectively.

@@

Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed and implemented in the Matita system to address exactly this problem. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution (described in the next section) that avoids the problem altogether by using predicate-like type classes in place of bundled records.

@@

The observant reader may wonder whether it might be beneficial to go one step further and unbundle proofs of laws and inherited substructures as well. This is not the case, because there is no point in sharing them. After all, by (weak) proof irrelevance, the ``value'' of such proofs can be of no consequence anyway. Indeed, parameterizing on proofs would be actively harmful because instantiations differing only in the proof argument would express the same thing yet be nonconvertible, requiring awkward conversions and hindering automation.

@@

Without further support from the system, this would have to be written as
\begin{lstlisting}
  Theorem L G e op {P: SemiGroup G e op}:
    Π x y z, e (op (op x y) z) (op (op z y) x).
\end{lstlisting}
Because \lstinline|e| and \lstinline{op} are freshly introduced local names, we cannot bind notations to them prior to this theorem. Hence, if we want notations, what we really need are canonical names for these components. This is easily accomplished with single-field type classes containing one component each, which we will call \emph{operational type classes}\footnote{These single-field type classes are used in the same way in the \lstinline|Clean| standard library~\cite{clean}.}:
\begin{lstlisting}
  Class Equiv A := equiv: relation A.
  Class SemiGroupOp A := sg_op: A $\to$ A $\to$ A.

@@

\begin{lstlisting}
  Theorem L G e op {P: SemiGroup G e op}: Π x y z,
    equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op z y) x).
\end{lstlisting}
where e and the op's are filled in by instance resolution.

@@

While we are on the topic of implicit generalization, we mention one inadequacy concerning their current implementation that we feel should be addressed for the facility to be a completely satisfying solution. While the syntax already supports variants (not shown above) that differ in how exactly different kinds of arguments are inferred and/or generalized, there is no support to have an argument be ``inferred if possible, generalized otherwise". We have found that the need for such a policy arises naturally when declaring a parameter of class type in a context where \emph{some} of its components are already available, while others are to be newly introduced. The current workaround in these cases involves providing names for components that are then never referred to, which is a bit awkward.

@@

Before we proceed with a discussion of the algebraic hierarchy based on predicate classes and operational classes, we briefly highlight one specific operational type class because we will use it later and because it is a particularly nice illustration of another neat application of operational type classes. The operation in question is that of deciding a proposition:
\begin{lstlisting}
  Class Decision (P: Prop): Type := decide: sumbool P (¬ P).
\end{lstlisting}
Here, \lstinline|sumbool| is just the (informative) sum of proofs.

@@

Fields declared with \lstinline|:>| are registered as hints for instance resolution, so that in any context where \lstinline|(A, =, +, 0, *, 1)| is known to be a \lstinline|SemiRing|, \lstinline|(A, =, +, 0)| and \lstinline|(A, =, *, 1)| are automatically known to be \lstinline|CommutativeMonoid|s (and so on, transitively, because instance resolution is recursive). In our hierarchy, these ``substructures'' by themselves establish the following inheritance diagram:

\vspace{5mm}
\begin{center}
\includegraphics{hierarchy.pdf}
\end{center}
\vspace{5mm}

@@

Having described the basic principles of our approach, in the remainder of this paper we tour other parts of our development, further illustrating what a state of the art formal development of foundational mathematical structures can look like with a modern proof assistant based on type theory.

@@

Initiality, too, is defined by a combination of an operational and a predicate class:
\begin{lstlisting}
  Context `{Category X}.
  Class InitialArrows (x: X): Type := initial_arrow: Π y, x ⟶ y.
  Class Initial (x: X) `{InitialArrows x}: Prop :=
    initial_arrow_unique: Π y (a: x ⟶ y), a = initial_arrow y.
\end{lstlisting}
The operational class \lstinline|InitialArrows| designates the arrows that originate from an initial object \lstinline|x| by virtue of it being initial. The \lstinline|Initial| class itself further requires these ``initial arrows'' to be unique. Having \lstinline|InitialArrows| as an operational type class means that we can always simply say \lstinline|initial_arrow y| whenever \lstinline|y| is known to be an object in a category known to have an initial object (where ``known'' should be read as ``can be determined by instance resolution'').

@@

  Class Functor `{Fmap}: Prop :=
    { functor_from: Category C
    ; functor_to: Category D
    ; functor_morphism:> Π a b: C, Setoid_Morphism (@fmap _ a b)
    ; preserves_id: `(fmap (cat_id: a ⟶ a) = cat_id)
    ; preserves_comp `(f: y ⟶ z) `(g: x ⟶ y):
      fmap (f ◎$$ g) = fmap f ◎$$ fmap g }.
\end{lstlisting}
We ought to say a few words about our use of \lstinline|fmap|. The usual mathematical notational convention for functor application is to use the name of the functor to refer to both its object map and its arrow map, relying on additional conventions regarding object/arrow names for disambiguation: ``\lstinline|F x|'' and ``\lstinline|F f|'' map an object and an arrow, respectively, because ``\lstinline|x|'' and ``\lstinline|f|'' are conventional names for objects and arrows, respectively.

In Coq, for a term \lstinline|F| to function as though it had two different types simultaneously (namely the object map and the arrow map), there must either (1) be coercions from the type of F to either function, or (2) F must be (coercible to) a single function that is able to consume both object and arrow arguments. In addition to not being supported by Coq, option (1) would violate our policy of leaving components unbundled. For (2), if it could be made to work at all (which is not clear at all), F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map (namely \lstinline|Π x y, (x ⟶ y) → (F x ⟶ F y)|) must refer to the object map.

We feel that these issues are not limitations of the Coq system, but merely reflect the fact that notationally identifying these two distinct and interdependent maps is an abuse of notation of sufficient severity to make it ill-suited to a formal development where software engineering concerns apply. Hence, we do not adopt this practice, and use ``\lstinline|fmap F|'' (name taken from the Haskell standard library) to refer to the arrow map of a functor \lstinline|F|.

@@

To specify the natural numbers and the integers as initial objects in the categories of semirings and rings, respectively, definitions of these categories are needed. While one could define both of them ``manually'', greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, for which \emph{varieties} can be defined generically (varieties are categories consisting of models for a fixed theory with homomorphisms between them).

@@

  Class AlgebraOps (σ: Signature) (carriers: sorts σ$$ → Type)
    := algebra_op: Π o: operation σ, op_type carriers (σ o).
\end{lstlisting}
Because our carriers will normally be equipped with a setoid equality, we further define the predicate class \lstinline|Algebra|, stating that each of the operations respects the setoid equality on the carriers:
\begin{lstlisting}
  Class Algebra `{Π a, Equiv (carriers a)} `{AlgebraOps σ$$ carriers}: Prop :=
      { algebra_setoids:> Π a, Setoid (carriers a)
      ; algebra_propers:> Π o: σ, Proper (=) (algebra_op o) }.
\end{lstlisting}
The \lstinline|(=)| referred to in \lstinline|algebra_propers| is an \lstinline|Equiv| instance expressing setoid-respecting extensionality for the function types generated by \lstinline|op_type|.

@@

To characterize structures such as semirings and rings, we need equational theories, consisting of a signature together with laws (represented by a predicate over equality entailments):
\begin{lstlisting}
  Record EquationalTheory :=
    { et_sig:> Signature
    ; et_laws:> EqEntailment et_sig → Prop }.
\end{lstlisting}
An \lstinline|EqEntailment| consists of premises and a conclusion represented by an inductively defined statement grammar, which in turn uses an inductively defined term grammar. A detailed discussion of these definitions and the theory developed for them is beyond the scope of this paper.

We now introduce a predicate class designating algebras that satisfy the laws of an equational theory:
\begin{lstlisting}
  Class InVariety
    (et: EquationalTheory) (carriers: sorts et → Type)
    {e: Π a, Equiv (carriers a)} `{!AlgebraOps et carriers}: Prop :=
    { variety_algebra:> Algebra et carriers
    ; variety_laws: Π s, et_laws et s → (Π $\hspace{-1.5mm}$vars, eval_stmt et vars s) }.
\end{lstlisting}
  % todo: why do i need the manual \hspace!?

@@

In addition to this variety category, we also have categories of lawless algebras, as well as forgetful functors from the former to the latter, and from the latter to the category of setoids.

@@

To give a further taste of what universal algebra in our development looks like, we consider the definitions involved in the first homomorphism theorem~\cite{meinke1993universal} in more detail:
\begin{theorem}[First homomorphism theorem]
If $A$ and $B$ are algebras, and $f$ is a homomorphism from $A$ to $B$, then the equivalence relation defined by $a\sim b$ if and only if $f(a)=f(b)$ is a congruence on $A$, and the algebra $A/\hspace{-1mm}\sim$ is isomorphic to the image of $f$, which is a subalgebra of B.
\end{theorem}

A set of relations \lstinline|e| (one for each sort) is a congruence for an existing algebra if (1) \lstinline|e| respects that algebra's existing setoid equality, and (2) the operations with \lstinline|e| again form an algebra (namely the quotient algebra):
\begin{lstlisting}
Context `{Algebra σ$$ A}.
Class Congruence (e: Π s: sorts σ, relation (v s)): Prop :=
  { congruence_proper:> Π s, Proper (equiv ==> equiv ==> iff) (e s)
  ; congruence_quotient:> Algebra σ$$ v (e:=e) }.
\end{lstlisting}
We have proved that this natural and economical type-theoretic formulation that leverages our systematic integration of setoid equality is equivalent to the traditional definition of congruences as relations that, represented as sets of pairs, form a subalgebra of the product algebra.

@@

With these instances in place, we can now define the improved natural numbers specification:
\begin{lstlisting}
  Context `{SemiRing A} `{!NaturalsToSemiRing A}.
  Class Naturals: Prop :=
    { naturals_ring:> SemiRing A
    ; naturals_to_semiring_mor:> Π `{SemiRing B},
        SemiRing_Morphism (naturals_to_semiring A B)
    ; naturals_initial:> Initial (bundle_semiring A) }.
\end{lstlisting}
Basing theory and programs on this abstract interface instead of on specific implementation (such as the ubiquitous Peano naturals \lstinline|nat| in the Coq standard library) is not only cleaner mathematically, but also facilitates easy swapping between implementations. And this benefit is far from theoretical, as diverse representations of the natural numbers are abound; for instance, unary, binary, factor multisets, and arrays of native machine words.

@@

To permit a \emph{generic} function operating on naturals to take advantage of specialized operations, we simply introduce an additional instance parameter:
\begin{lstlisting}
Definition calculate_things `{Naturals N} `{Π$$n m: N, Decision (n = m)}
  (a b: nat): ... := ... decide (a = b) ... .
\end{lstlisting}
Without the \lstinline|Decision| parameter \lstinline|calculate_things| would be equally correct, but could be less efficient. Thus, by this scheme one can start by writing correct-but-possibly-inefficient programs that make use of generic operation instances, and then selectively improve efficiency of key algorithms simply by adding additional operational type class instance parameters where profiling shows it to make a significant difference, without changing their definition body.

@@

In our development we prove that the standard library's default rationals do indeed implement \lstinline|Rationals|, as do implementations of the \lstinline|QType| module interface. While the latter is rather ad-hoc from a theoretical perspective, it is nevertheless of great practical interest because it is used for the very efficient \lstinline|BigQ| rationals based on machine integers~\cite{machineintegers}. Hence, theory and programs developed on our \lstinline|Rationals| interface applies and can make immediate use of these efficient rationals. We plan to rebase the computable real number implementation in CoRN on this interface, precisely so that it may be instantiated with efficient implementations like these.

We also plan to provide an abstract interface for polynomials as a free commutative algebra. This would unify existing implementations such as coefficient lists and Bernstein polynomials.

@@

Traditionally, proof assistants such as Coq have provided sophisticated tactics or built-in commands to support this ``quoting''. In~\cite{Hints}, Asperti et. al. show that unification hints, a very general way of facilitating user-defined extensions to term and type inference, can be used to semi-automatically build quote functions without dropping to a meta-level.\footnote{Gonthier provides similar functionality by ingeniously using canonical structures.} This feature is absent from Coq, but fortunately type classes also allow us to do this, as we will now show.

@@

With these, we can define the primary ingredient, the \lstinline|Quote| class:
\begin{lstlisting}
  Class Quote {V} (l: Vars V) (n: Value) {V'} (r: Vars V'): Type :=
    { quote: Expr (V + V')
    ; eval_quote: eval (merge l r) quote = n }.
\end{lstlisting}
We can think of \lstinline|Quote| as the type for a family of Prolog-like syntax-directed resolution functions, which will take as input \lstinline|V| and \lstinline|l| representing previously encountered holes (opaque subexpressions that could not be destructured further) and their values, along with a concrete term \lstinline|n| to be quoted. Their ``output'' will consist not only of the fields in the class, but also of \lstinline|V'| and \lstinline|r| representing additional holes and their values. Hence, a type class constraint of the form \lstinline|Quote x y z| should be read as ``quoting \lstinline|y| with existing heap \lstinline|x| generates new heap \lstinline|z|''.

@@

Using a small additional utility lemma we can also easily quote equalities into \lstinline|eval p e = eval p e'| form with a shared variable pack (so that identical expressions on both sides of the equation don't end up as different variables).

@@

Mathematically, finite sequences can be characterized as free monoids over sets. A categorical way of expressing this is in terms of adjunctions. As with the numeric interfaces, we \emph{could} fully embrace this perspective, paying no heed to practicality of implementation and usage, and define a relatively succinct type class for sequences as follows:
\begin{lstlisting}
  Class PoshSequence
    (free: setoid.Object → monoid.Object) `{Fmap free}
    (singleton: id ⇛ monoid.forget ∘ free)
    (extend: `((x ⟶ monoid.forget y) → (free x ⟶ y))): Prop :=
    { sequence_adjunction: AltAdjunction singleton extend
    ; extend_morphism: `(Setoid_Morphism (extend x y)) }.
\end{lstlisting}
Here, \lstinline|monoid.forget| is the forgetful functor from monoids to sets.

@@

We emphasize that these efficiency problems only affect type checking; efficiency of computation using type-checked terms is not affected.
